The function--sheaf dictionary for smooth commutative group schemes over finite fields with finitely generated geometric component group

### 2014-01-24 dictionary.tex
More use of `functorial isomorphism' in dictionary.

\begin{abstract}
  We extend Deligne's function--sheaf dictionary from
  connected commutative algebraic groups over finite fields to smooth
  commutative group schemes locally of finite type over finite fields
  with finitely-generated geometric component group.
  To do so, we introduce the category of quasicharacter sheaves on such group schemes
  $G$ over $\mathbb{F}_q$ and exhibit a canonical isomorphism between the character group of $G(\mathbb{F}_q)$ and the group of isomorphism classes of quasicharacter sheaves on $G$.
   \end{abstract}

@@

Although hidden in the discussion above, the connectness of $G$ played a crucial role in the geometrization and categorification of characters of $G(\Fq)$.  So it is natural to ask if the sheaf--function dictionary can be generalized to not-necessarily connected commutative algebraic groups over finite fields. In this paper we answer that question. 
%
In fact, we show much more by exhibiting a sheaf--function dictionary, in the sense above, for any smooth commutative group scheme $G$, locally of finite type over a finite field $\Fq$, 
for which the geometric component group of $G$ is finitely generated; we call such group schemes $\QC$-schemes.
To do this we must pass from $\ell$-adic local systems on a $\QC$-scheme $G$ to Weil local systems on $\bG$ 
and also keep track of the isomorphisms \eqref{introbox} (existence of such an isomorphism is not enough), leading to the notion of {\it quasicharacter sheaves on $G$}.

@@

A simple example will illustrate some of the issues that arise when extending the sheaf--function dictionary to $\QC$-schemes. Consider the etale group scheme $G = \ZZ_{/\Fq}$. Characters of $G(\Fq) = \ZZ$ are neither finite nor bounded, in general. For example, the $\ell$-adic character $\ZZ \to \EEx$ defined by $1 \mapsto 1+ \ell$ is bounded but not finite, while the character $\ZZ \to \EEx$ defined by $1 \mapsto \ell$ is unbounded.  As we show in the paper, $\ell$-adic local systems on $G$ cannot be used to geometrize either of these characters; however, quasicharacter sheaves do the job, handily. We remark that such characters arise when considering unramified quasicharacters of the $p$-adic group $\Qp^\times$. This example informs our choice of the term ``quasicharacter sheaf''. Indeed, in \cite{cunningham-roe:p-adic_tori} we use this paper to geometrize and categorify all quasicharacters of all algebraic tori over all local fields.

@@

We close the introduction with a more detailed description of the structure of this paper.
After setting some notation in Section~\ref{sec:QCschemes}, in Section~\ref{sec:category} we define quasicharacter sheaves and give an interpretation of
them as Weil sheaves with certain properties.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism,
one direction of the function--sheaf dictionary.
We define pullbacks and products of quasicharacter sheaves in Section~\ref{sec:pullback}, and use
them to prove a functoriality result on the trace of Frobenius.
In Section~\ref{sec:basechange} we study how quasicharacter sheaves behave under base change
and Weil restriction.
Sections~\ref{sec:bounded} and \ref{sec:finite} describe bounded quasicharacter sheaves and finite quasicharacter sheaves, respectively,
forming two important subcategories of $\QC(G)$.  These subcategories play a key
role in the proof of Proposition~\ref{prop:restriction}.
In section \ref{sec:connected} we show that bounded and finite quasicharacter sheaves are the same
as standard quasicharacter sheaves when $G$ is connected, and that trace of Frobenius is an isomorphism.
We shift to \'etale group schemes $G$ in Section~\ref{sec:etale} and give a more concrete interpretation
of quasicharacter sheaves in this case.  Using a notion of global sections for quasicharacter sheaves,
we give isomorphisms between $\QCiso{G}$, $\Hh^1(\Weil{}, \Hom(G(\bFq), \EEx))$ and $\Hom(G(\Fq), \EEx)$, where $\Weil{}$ is the Weil group for $\Fq$.
In Section~\ref{sec:restriction} we show that for non-connected $G$, restriction to the identity component
induces an essentially surjective functor of quasicharacter sheaves.
Section~\ref{sec:snake} presents the main theorem of this paper: trace of Frobenius
is an isomorphism for $G$ as long as the geometric component group is finitely generated.

@@

\begin{theorem}\label{thm:snake}
  If $G$ is a smooth, commutative group scheme, locally of finite type over
  $\Fq$, and if the abelian group $\pi_0(G)(\bFq)$
  is finitely generated, then the trace of Frobenius
  \[
  \trFrob{G} : \QCiso{G} \to G(\Fq)^*
  \]
  is an isomorphism of groups.
\end{theorem}

@@

With Corollary~\ref{cor:bounded-and-finite} in hand 
we can see why neither finite quasicharacter sheaves nor bounded quasicharacter sheaves 
are up to the task of geometrizing all characters of $G(\Fq)$ for general $\QC$-schemes $G$ since, in general, neither of the full subcategories
$\QCf(G) \subset \QCb(G) \subset \QC(G)$ are equivalences.
For example, consider the case when $G$ is the discrete \'etale group scheme $\ZZ$.
If $\chi : \ZZ \to \EEx$ is the quasicharacter of $G(\Fq)$ determined by $\chi(1) = \ell$ 
and if $\qcs{L}$ is a quasicharacter sheaf on $G$ in the isomorphism class 
corresponding to $\chi$ under Theorem~\ref{thm:snake},
then $\qcs{L}$ is not a bounded quasicharacter sheaf since the image of $\chi$ is not
bounded.  
On the other hand, if $\chi : \ZZ \to \EEx$ is the quasicharacter of $G(\Fq)$ determined by $\chi(1) = 1+\ell$ 
and if $\qcs{L}$ is a quasicharacter sheaf on $G$ in the isomorphism class 
corresponding to $\chi$ under Theorem~\ref{thm:snake},
then $\qcs{L}$ is a bounded quasicharacter sheaf but not a finite quasicharacter sheaf
since the image of $\chi$ is bounded but not finite.
In this latter case the stalk of $\qcs{L}$ at the component $1$ is defined
by a pro-\'etale cover rather than an \'etale cover.

>>>>

\begin{abstract}
  We extend Deligne's function--sheaf dictionary from
  connected commutative algebraic groups over finite fields to smooth
  commutative group schemes locally of finite type over finite fields
  with finitely-generated geometric component group.
  To do so, we introduce the category of quasicharacter sheaves on such group schemes
  $G$ over $\mathbb{F}_q$ and exhibit a functorial isomorphism between the character group of $G(\mathbb{F}_q)$ and the group of isomorphism classes of quasicharacter sheaves on $G$.
   \end{abstract}

@@

Although hidden in the discussion above, the connectness of $G$ played a crucial role in the geometrization and categorification of characters of $G(\Fq)$.  So it is natural to ask if the sheaf--function dictionary can be generalized to not-necessarily connected commutative algebraic groups over finite fields. In this paper we answer that question. 
%
In fact, we show much more by exhibiting a sheaf--function dictionary, in the sense above, for any smooth commutative group scheme $G$, locally of finite type over a finite field $\Fq$, 
for which the geometric component group of $G$ is finitely generated; we call such group schemes $\QC$-schemes.
To do this we must pass from $\ell$-adic local systems on a $\QC$-scheme $G$ to Weil local systems on $G$ 
and also keep track of the isomorphisms \eqref{introbox} (existence of such an isomorphism is not enough), leading to the notion of {\it quasicharacter sheaves on $G$}.

@@

A simple example will illustrate some of the issues that arise when extending the sheaf--function dictionary to $\QC$-schemes. Consider the etale group scheme $G = \ZZ_{/\Fq}$. Characters of $G(\Fq) = \ZZ$ are neither finite nor bounded, in general. For example, the $\ell$-adic character $\ZZ \to \EEx$ defined by $1 \mapsto 1+ \ell$ is bounded but not finite, while the character $\ZZ \to \EEx$ defined by $1 \mapsto \ell$ is unbounded.  As we show in the paper, $\ell$-adic local systems on $G$ cannot be used to geometrize either of these characters; however, quasicharacter sheaves do the job, handily. We remark that such characters arise when considering unramified quasicharacters of the $p$-adic group $\Qp^\times$  This example informs our choice of the term ``quasicharacter sheaf''. Indeed, in \cite{cunningham-roe:p-adic_tori} we use this paper to geometrize and categorify all quasicharacters of all algebraic tori over all local fields.

@@

We close the introduction with a more detailed description of the structure of this paper.
After setting some notation in Section~\ref{sec:QCschemes}, in Section~\ref{sec:category} we define quasicharacter sheaves and give an interpretation of
them as Weil sheaves with certain properties.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism,
one direction of the function--sheaf dictionary.
We define pullbacks and products of quasicharacter sheaves in Section~\ref{sec:pullback}, and use
them to prove a functoriality result on the trace of Frobenius.
In Section~\ref{sec:basechange} we study how quasicharacter sheaves behave under base change
and Weil restriction.
Sections~\ref{sec:bounded} and \ref{sec:finite} describe bounded quasicharacter sheaves and finite quasicharacter sheaves, respectively,
forming two important subcategories of $\QC(G)$.  These subcategories play a key
role in the proof of Proposition~\ref{prop:restriction}.
In section \ref{sec:connected} we show that bounded and finite quasicharacter sheaves are the same
as standard quasicharacter sheaves when $G$ is connected, and that trace of Frobenius is an isomorphism.
We shift to \'etale group schemes $G$ in Section~\ref{sec:etale} and give a more concrete interpretation
of quasicharacter sheaves in this case.  Using a notion of global sections for quasicharacter sheaves,
we give isomorphisms between $\QCiso{G}$, $\Hh^1(\Weil{}, \Hom(G(\bFq), \EEx))$ and $\Hom(G(\Fq), \EEx)$, where $\Weil{}$ is the Weil group for $\Fq$.
In Section~\ref{sec:restriction} we show that for non-connected $G$, restriction to the identity component
induces an essentially surjective functor of quasicharacter sheaves.
Section~\ref{sec:snake} presents the main theorem of this paper: trace of Frobenius
is a functorial isomorphism for $\QC$-schemes.

@@

\begin{theorem}\label{thm:snake}
  If $G$ is a smooth, commutative group scheme, locally of finite type over
  $\Fq$, and if the abelian group $\pi_0(G)(\bFq)$
  is finitely generated, then the trace of Frobenius
  \[
  \trFrob{G} : \QCiso{G} \to G(\Fq)^*
  \]
  is a functorial isomorphism of groups. 
\end{theorem}

@@

With Corollary~\ref{cor:bounded-and-finite} in hand 
we can see why neither finite quasicharacter sheaves nor bounded quasicharacter sheaves 
are up to the task of geometrizing all characters of $G(\Fq)$ for general $\QC$-schemes $G$ since, in general, neither of the full subcategories
$\QCf(G) \subset \QCb(G) \subset \QC(G)$ are equivalences.
For example, consider the case when $G$ is the discrete \'etale group scheme $\ZZ$.
If $\chi : \ZZ \to \EEx$ is the quasicharacter of $G(\Fq)$ determined by $\chi(1) = \ell$ 
and if $\qcs{L}$ is a quasicharacter sheaf on $G$ in the isomorphism class 
corresponding to $\chi$ under Theorem~\ref{thm:snake},
then $\qcs{L}$ is not a bounded quasicharacter sheaf since the image of $\chi$ is not
bounded.  
Likewise, if $\chi : \ZZ \to \EEx$ is the quasicharacter of $G(\Fq)$ determined by $\chi(1) = 1+\ell$ 
and if $\qcs{L}$ is a quasicharacter sheaf on $G$ in the isomorphism class 
corresponding to $\chi$ under Theorem~\ref{thm:snake},
then $\qcs{L}$ is a bounded quasicharacter sheaf but not a finite quasicharacter sheaf
since the image of $\chi$ is bounded but not finite.
In this latter case the stalk of $\qcs{L}$ at the component $1$ is defined
by a pro-\'etale cover rather than an \'etale cover.
